Covariates analysis of MSBB reprocessed RNASeq data

### 2016-06-01 code/Rmd/MSSM_RNASeq_cdr_Reprocessed.Rmd
Major update on the reprocessed RNASeq processing code for MSSM

cl = makeCluster(2)
registerDoParallel(cl)

@@

Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn6114414'
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, quote = '', header = T, sep = '\t')

@@

METADATA = left_join(METADATA_TECHNICAL %>%
                       dplyr::mutate(ID = gsub('hB_RNA_','',NewBarcode),
                                     ID = as.numeric(ID)),
                     METADATA_CLINICAL %>%
                       dplyr::select(BB, BM.44.Barcode, BM36.Fnl, BM22.Fnl, BM10.Fnl, PMI, RACE, 
                                     AOD, CDR, SEX, NP.1, PlaqueMean, bbscore) %>%
                       dplyr::rename(BM_44 = BM.44.Barcode,
                                     BM_10 = BM10.Fnl,
                                     BM_22 = BM22.Fnl,
                                     BM_36 = BM36.Fnl) %>%
                       tidyr::gather(BrainRegion, ID, -BB, -PMI, -RACE, 
                                     -AOD, -CDR, -SEX, -NP.1, -PlaqueMean, -bbscore) %>%
                       filter(!is.na(ID)) %>%
                       dplyr::mutate(ID = as.numeric(ID)))
rownames(METADATA) = METADATA$OldBarcode

# Filter data
ind <- intersect(rownames(METADATA), colnames(COUNT))
# ind <- setdiff(ind, 'hB_RNA_10622')
METADATA <- METADATA[ind,]
COUNT <- COUNT[, ind]
```
### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
# Max out age at 90
METADATA$AOD = gsub('\\+','',as.character(METADATA$AOD))

@@

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$OldBarcode

@@

# Sex
vcd::mosaic(~ SEX + CDR + BrainRegion, data = METADATA)
```

### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,'NULL design(voom-normalized)', isKeyPlot=TRUE, MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

@@

@@

# Store differential expression results
write.table(DE, file = 'MSSMdiffExpression.tsv', sep = '\t', row.names=F, quote=F)
dexp_obj = File('MSSMdiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
dexp_obj = synStore(dexp_obj, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                    |
|  -------                  |   ---------                     |
|  Covariates               |  `r covar_obj$properties$id`    |
|  Residual Expression      |  `r expr_obj$properties$id`     |
|  Differential Expression  |  `r dexp_obj$properties$id`     |

>>>>

cl = makeCluster(14)
registerDoParallel(cl)

@@

Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_B1_ID <- 'syn6114414'
ALL_USED_IDs <- COUNT_B1_ID
COUNT_B1 <- read.table(synGet(COUNT_B1_ID)@filePath, quote = '', header = T, sep = '\t')

COUNT_B2_3_ID <- 'syn6125203'
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_B2_3_ID
COUNT_B2_3 <- read.table(synGet(COUNT_B2_3_ID)@filePath, quote = '', header = T, sep = '\t')

# Download expression data
COUNT_B4_ID <- 'syn6130073'
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_B4_ID
COUNT_B4 <- read.table(synGet(COUNT_B4_ID)@filePath, quote = '', header = T, sep = '\t')

# Remove samples in Batch 2&3 if they present in batch 4
COUNT_B2_3 = COUNT_B2_3[,setdiff(colnames(COUNT_B2_3), colnames(COUNT_B4))]

COUNT = cbind(COUNT_B1, COUNT_B2_3[rownames(COUNT_B1),], COUNT_B4[rownames(COUNT_B1),])

# Get sample ids
SampleID = data.frame(SampleID = colnames(COUNT),
                 ID = colnames(COUNT)) %>% 
  tidyr::separate(ID, c('A','B','ID'), sep = '_') 
SampleID$ID[SampleID$ID == 'HISEQ'] = SampleID$B[SampleID$ID == 'HISEQ']

SampleID = SampleID %>%
  dplyr::select(SampleID, ID) %>%
  dplyr::mutate(ID = as.numeric(ID))

@@

METADATA = left_join(METADATA_TECHNICAL %>%
                       dplyr::mutate(ID = gsub('hB_RNA_','',NewBarcode),
                                     ID = as.numeric(ID)),
                     METADATA_CLINICAL %>%
                       dplyr::select(BB, BM.44.Barcode, BM36.Fnl, BM22.Fnl, BM10.Fnl, PMI, RACE, 
                                     AOD, CDR, SEX, NP.1, PlaqueMean, bbscore) %>%
                       dplyr::rename(BM_44 = BM.44.Barcode,
                                     BM_10 = BM10.Fnl,
                                     BM_22 = BM22.Fnl,
                                     BM_36 = BM36.Fnl) %>%
                       tidyr::gather(BrainRegion, ID, -BB, -PMI, -RACE, 
                                     -AOD, -CDR, -SEX, -NP.1, -PlaqueMean, -bbscore) %>%
                       filter(!is.na(ID)) %>%
                       dplyr::mutate(ID = as.numeric(ID))) %>%
  left_join(SampleID) %>%
  filter(!is.na(SampleID)) %>%
  unique
rownames(METADATA) = METADATA$SampleID

# Filter data
ind <- intersect(METADATA$SampleID, colnames(COUNT))
indRemoved <- setdiff(colnames(COUNT), METADATA$SampleID)
# ind <- setdiff(ind, 'hB_RNA_10622')
METADATA <- METADATA[ind,]
COUNT <- COUNT[, ind]
```
### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
# Max out age at 90
METADATA$AOD = gsub('\\+','',as.character(METADATA$AOD))

@@

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$SampleID

@@

# Sex
vcd::mosaic(~ SEX + CDR + BrainRegion, data = METADATA)
```
### Quality Control
```{r check.sex}
# Query biomart for Y chromosome genes
ensembl = useMart('ensembl', dataset = 'hsapiens_gene_ensembl', host = 'www.ensembl.org')
EnsembleInY <- getBM(attributes = c("ensembl_gene_id"),
                       filters = "chromosome_name", 
                       values = 'Y',
                       mart = ensembl)

YChromGenes = data.frame(Gene.ID = rownames(COUNT),
                             ID = rownames(COUNT)) %>%
  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.') %>%
  inner_join(EnsembleInY)

tmp1 = t(COUNT[YChromGenes$Gene.ID,]) %>%
  rowSums %>% rownameToFirstColumn('SampleID') %>%
  dplyr::rename(CPM.Sum = DF) %>%
  left_join(COVARIATES %>% 
              rownameToFirstColumn('SampleID'))

ggplot(tmp1, aes(x = SEX, y= log2(CPM.Sum))) + geom_boxplot()

EnsembleInX <- getBM(attributes = c("ensembl_gene_id"),
                       filters = "chromosome_name", 
                       values = 'X',
                       mart = ensembl)

XChromGenes = data.frame(Gene.ID = rownames(COUNT),
                             ID = rownames(COUNT)) %>%
  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.') %>%
  inner_join(EnsembleInX)

tmp1 = t(COUNT[XChromGenes$Gene.ID,]) %>%
  rowSums %>% rownameToFirstColumn('SampleID') %>%
  dplyr::rename(CPM.Sum = DF) %>%
  left_join(COVARIATES %>% 
              rownameToFirstColumn('SampleID'))

ggplot(tmp1, aes(x = SEX, y= log2(CPM.Sum))) + geom_boxplot()
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

@@

# Design Matrix
write.table(DM1$design, file='MSSMDesignMat.tsv', sep = '\t', row.names=F, quote=F)
obj = File('MSSMDesignMat.tsv', 
           name = 'Design Matrix',
           parentId = CODE$properties$id)
dm_obj = synStore(obj,
                    used = ALL_USED_IDs, 
                    activityName = activityName, 
                    executed = thisFile, 
                    activityDescription = activityDescription)

@@

# Store differential expression results
write.table(DE, file = 'MSSMdiffExpression.tsv', sep = '\t', row.names=F, quote=F)
dexp_obj = File('MSSMdiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
dexp_obj = synStore(dexp_obj, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                    |
|  -------                  |   ---------                     |
|  Covariates               |  `r covar_obj$properties$id`    |
|  Raw counts (log CPM)     |  `r rawcpm_obj$properties$id`   |
|  Design Matrix            |  `r dm_obj$properties$id`       |
|  Residual Expression      |  `r expr_obj$properties$id`     |
|  Differential Expression  |  `r dexp_obj$properties$id`     |
